Type theory with single substitutions\thanks{This research was supported by the HOTT ERC Grant 101170308.}

### 2025-07-01 lfmtp/p.tex
stuff

\title{Type theory with single substitutions\thanks{This research was supported by the the HOTT ERC Grant 101170308.}}
\author{Ambrus Kaposi
\institute{Eötvös Loránd University (ELTE)\\
Budapest, Hungary}
\email{akaposi@inf.elte.hu}
\and
Szumi Xie
\institute{Eötvös Loránd University (ELTE)\\
Budapest, Hungary}
\email{szumi@inf.elte.hu}
}
\def\titlerunning{Type theory with single substitutions}
\def\authorrunning{A. Kaposi \& Sz. Xie}

@@

What is type theory? Here we refer to type theory as a particular
formal system based on Martin-Löf's original definition
\cite{martinlof73predicative}, and not to the study of type systems
(e.g.\ \cite{DBLP:books/daglib/0005958}).

@@

However, even at the GAT level of abstraction, there are some choices
to be made about the instantiation operation. It is usually convenient
to use parallel substitutions, which means that the instantiation
operation replaces all variables in a context by terms in another
context \emph{at the same time}, that is, $t[x_0\mapsto
  u_0,\dots,x_n\mapsto u_n]$ where $x_0,\dots,x_n$ are the variables
in the context of $t$. We call $(x_0\mapsto u_0,\dots,x_n\mapsto u_n)$
a parallel substitution, where the terms $u_0,\dots,u_n$ are
all in the same context. Parallel substitutions are very natural in
the algebraic setting because they form the morphisms in a category
where objects are contexts (lists of types), then types/terms are
presheaves, context extension is a representability condition, so the
equations of the algebraic theory are forced by the categorical
structure. Some examples of algebraic definitions of type theory using
parallel substitutions ranging from the more syntactic to the more
semantic: Ehrhard's notion \cite{ehrhard,coquandEhrhard}, categories
with families (CwF \cite{DBLP:conf/types/Dybjer95,Castellan2021}),
natural models \cite{DBLP:journals/mscs/Awodey18}, contextual
categories \cite{DBLP:journals/apal/Cartmell86}, C-systems
\cite{DBLP:journals/lmcs/AhrensLV18}, locally cartesian closed
categories \cite{DBLP:journals/mscs/ClairambaultD14}, path categories
\cite{10.1145/3204492}.

@@

\vspace{-1em}
\subparagraph*{Formalisations of type theory.}
Most computer formalisations of type theory are extrinsic
\cite{DBLP:journals/pacmpl/0001OV18,DBLP:conf/cpp/AdjedjLMPP24,DBLP:journals/jar/SozeauABCFKMTW20}:
they define the syntax as abstract syntax trees, and equip it with
typing and conversion relations. A higher level representation is
working in setoid hell \cite{chapman09eatitself,setoidhell}: terms are
intrinsically well-typed, but conversion is still an explicit
relation. For formalising the GAT-level syntax, one needs a stronger
metatheory than plain Agda or Coq: the metatheory has to support
quotient inductive-inductive types (QIITs
\cite{DBLP:journals/pacmpl/KaposiKA19}), in other words, initial
models of GATs. Altenkirch and Kaposi
\cite{DBLP:conf/popl/AltenkirchK16} formalised the syntax of type
theory using postulated QIITs in Agda, together with parametricity and
normalisation \cite{lmcs:4005} proofs. Brunerie and de Boer
\cite{initiality-agda} constructed the initial contextual category in
Agda using postulated quotients. Although we don't have a general
proof, we have experimental evidence that QIITs are supported by the
cubical set model \cite{DBLP:conf/lics/CoquandHM18} and the setoid
model \cite{kaposi-qiit-setoid}, thus Cubical Agda
\cite{DBLP:journals/jfp/VezzosiMA21} and setoid/observational type
theory \cite{setoid,DBLP:phd/hal/Pujet22} support QIITs. For example,
in Cubical Agda, set-truncated and groupoid-truncated syntaxes of type
theory have been formalised \cite{cohtt}. Without QIITs, tricks such
as shallowly embedding the syntax can be used to formalise metatheoretic
results about type theory such as gluing
\cite{glue}, and its special cases
canonicity and parametricity \cite{kaposi-shallow}.

@@

First of all, we need a sort of terms which have to be indexed by
types because we only want well-typed terms. Both types and terms can
include variables, and to keep track of the currently available
variables, we also index them by contexts: a context is a list of
types, the length is the number of available variables, and we add new
variables at the end (that is, they are snoc-lists rather than
cons-lists). Types are also indexed by their universe level, this is a
technical requirement for avoiding Russell's paradox. The index $i$ is
an implicit argument of $\Tm$.
\begin{equation*}
\Con : \Set \hspace{3em} 
\Ty : \Con\ra\N\ra\Set \hspace{3em} 
\Tm : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set
\end{equation*}
Just as lists have two constructors, there are two ways to form
contexts: the empty context $\diamond$ and context extension $\ext$
which is like the snoc operation for lists. Context extension takes a
type which can have variables in the context preceding the type.
\[
\diamond : \Con \hspace{3em}
\blank\ext\blank : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con
\]
We will refer to variables by their distance from the end of the
context. We would like to describe the operation providing the last
variable in a context, first of all we need an operation providing the last
variable in the context. This is the zero De Bruijn index \cite{debruijn}, which we
denote by $\q : \Tm\,(\Gamma\ext A)\,A'$. It is a term in an
extended context, and it should have type $A$, but the issue is that
$A : \Ty\,\Gamma$, and $A' : \Ty\,(\Gamma\ext A)$. We need a way to
weaken the type $A$ to obtain such an $A'$. For this, we will
introduce an operation $\blank[\p] : \Ty\,\Gamma\,i\ra\Ty\,(\Gamma\ext
A)\,i$ and define $A' := A[\p]$. Instead of introducing just
$\blank[\p]$, we generalise a bit and add a new sort $\Sub$. For
now, $\Sub$ only has the single element $\p$ and we introduce the operation
$\blank[\blank]$ called \emph{instantiation}. Elements of $\Sub$
will later be called substitutions, hence the name, but for now, we
only have single end-of context weakenings in $\Sub$. The context
$\Delta$ is called the domain, $\Gamma$ the codomain of a
$\Sub\,\Delta\,\Gamma$.
\begin{alignat*}{10}
  & \Sub && : \Con\ra\Con\ra\Set &&\hspace{1.5em}
  & \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \hspace{1.5em}
  \p && : \Sub\,(\Gamma\ext A)\,\Gamma \hspace{1.5em}
  && \q && : \Tm\,(\Gamma\ext A)\,(A[\p])
\end{alignat*}
The operations $\p$ and $\q$ take three implicit parameters, $\Gamma$,
$i$ and $A$, while $\blank[\blank]$ takes $\Gamma$, $i$ and $\Delta$
implicitly. Still, we only have the last variable $\q$ in the context,
we don't have e.g.\ the last but one variable in $\Tm\,(\Gamma\ext
A\ext B)\,(A[\p][\p])$ where we had to weaken $A$ twice to make it fit
with its context. To obtain more variables, we also allow weakening of
terms, more precisely, we introduce an instantiation operation for
terms with an overloaded notation.
\begin{alignat*}{10}
& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma])
\end{alignat*}
Note that this is a dependent function as the type of the resulting
term has to be weakened the same way as the term itself. Now we can
define all variables counting from the end of the context (De Bruijn
indices): $0 := \q$, $1 := \q[\p]$, $2 := \q[\p][\p]$,
$3 := \q[\p][\p][\p]$, and so on.

@@

\cite{kaposi-shallow} we call it termification, as most sorts in the

@@

B-systems are B-frames
together with substitution, weakening and generic element
operations. A B-frame is given by sets $B_i$, $\tilde{B}_i$ and
functions between them as shown below . \\
\begin{tikzpicture}
  \node (b0) at (4,0) {$\top$};
  \node (b1) at (5,0) {$B_1$};
  \node (b2) at (12,0) {$B_2$};
  \node (b3) at (15,0) {$\dots$};
  \node (c1) at (7,1) {$\tilde{B}_1$};
  \node (c2) at (14,1) {$\tilde{B}_2$};
  \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{ft}_0$} (b0);
  \draw[->,font=\scriptsize] (b2) edge node[below] {$\mathsf{ft}_1$} (b1);
  \draw[->,font=\scriptsize] (b3) edge node[below] {$\mathsf{ft}_2$} (b2);
  \draw[->,font=\scriptsize] (c1) edge node[above] {$\partial_1$} (b1);
  \draw[->,font=\scriptsize] (c2) edge node[above] {$\partial_2$} (b2);
\end{tikzpicture} \\
Using our notation, this is the following data ($\diamond$ is the empty context, $\ext$ is context extension). \\
\begin{tikzpicture}
  \node (b0) at (4,0) {$\top$};
  \node (b1) at (5,0) {$\Ty\,\diamond$};
  \node (b2) at (12,0) {$(A:\Ty\,\diamond)\times\Ty\,(\diamond\ext A)$};
  \node (b3) at (15,0) {$\dots$};
  \node (c1) at (7,1)  {$(A:\Ty\,\diamond)\times\Tm\,\diamond\,A$};
  \node (c2) at (14,1) {$\big((A:\Ty\,\diamond)\times(B:\Ty\,(\diamond\ext A))\big)\times\Tm\,(\diamond\ext A)\,B$};
  \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{fst}$} (b0);
  \draw[->,font=\scriptsize] (b2) edge node[below] {$\mathsf{fst}$} (b1);
  \draw[->,font=\scriptsize] (b3) edge node[below] {$\mathsf{fst}$} (b2);
  \draw[->,font=\scriptsize] (c1) edge node[above] {$\mathsf{fst}$} (b1);
  \draw[->,font=\scriptsize] (c2) edge node[above] {$\mathsf{fst}$} (b2);
\end{tikzpicture} \\
The substitution operation $\mathbb{S}$ for an $x:\tilde{B}_{n+1}$ is
a homomorphism of the slice B-frames $\mathbb{B}/\partial(x) \ra
\mathbb{B}/\mathsf{ft}(\partial(x))$. In our notation, $x = (\Gamma,
A, a)$ where $a : \Tm\,\Gamma\,A$, and $\mathbb{S}$ corresponds to the
following functions for $\Ty$ (a map between the bottom rows of the
diagrams):
\begin{alignat*}{10}
  & \blank[\langle a\rangle] && : \Ty\,(\Gamma\ext A)\ra\Ty\,\Gamma \\
  & \blank[\langle a\rangle^+] && : \Ty\,(\Gamma\ext A\ext B)\ra\Ty\,(\Gamma\ext B[\langle a\rangle]) \\
  & \blank[\langle a\rangle^{++}] && : \Ty\,(\Gamma\ext A\ext B\ext C)\ra\Ty\,(\Gamma\ext B[\langle a\rangle]\ext C[\langle a\rangle^+]) \\
  & \dots,
\end{alignat*}
and similarly $\mathbb{S}$ also includes all the (lifted) substitution
operations for terms (top rows of the
diagrams). Analogously, the weakening operation corresponds to
$\blank[\p^{+\dots+}]$ operations, and the generic element is $\q$ in
our notation. The six groups of equations correspond to our
$[\langle\rangle][]$, $[\p][^+]$, $\q[^+]$, $[\p][\langle\rangle]$,
$\q[\langle\rangle]$, $[\p^+][\langle\q\rangle]$ equations, in this
order. However we don't include the lifted versions of these equations,
equations $[\langle\rangle][]$, $[\p^+][\langle\q\rangle]$ are only
stated for types, and $\q[^+]$, $\q[\langle\rangle]$ are only stated
for terms. The missing equations are admissible, see Section
\ref{sec:admissible}. In the presence of $\U$ and $\Pi$, we
reduce the needed equations even more, see Section
\ref{sec:minimisation}.

@@

We show that the roundtrip CwF $\longrightarrow$ SSC
$\xrightarrow{\text{termification}}$ CwF results in a contextually
isomorphic CwF. A contextual isomorphism
\cite{DBLP:journals/corr/abs-2211-07487} is a weak CwF-morphism
(pseudomorphism, i.e.\ context extension and the empty context are
only preserved up to isomorphism) which is bijective on types and
terms. Note that a contextual isomorphism preserves all type formers
specified by universal properties.
\begin{problem}
  Given a CwF (with type formers) $M$, let $M'$ denote the CwF
  obtained by first seeing it as an SSC model and then termifying it. We
  construct a contextual isomorphism between $M'$ and $M$.
\end{problem}
\begin{proof}[Construction]
  We denote the components of the contextual isomorphism $F$ as follows.
  \begin{alignat*}{10}
    & F : \Con_{M'}\ra\Con_M && F : \Ty_{M'}\,\Gamma \cong \Ty_M\,(F\,\Gamma) \\
    & F : \Sub_{M'}\,\Delta\,\Gamma\ra\Sub_M\,(F\,\Delta)\,(F\,\Gamma)\hspace{3em} && F : \Tm_{M'}\,\Gamma\,A \cong \Tm_M\,(F\,\Gamma)\,(F\,A)
  \end{alignat*}
  We define them in the same order, omitting the $M$ subscripts:
  $F\,\Gamma := \diamond\ext\Gamma$, 
  $F\,\gamma := (\p,\gamma[\p]\cdot\q)$, 
  $F\,A := \El\,(A[\p]\cdot\q)$, 
  $F\,a := a[\p]\cdot\q$.
%  \begin{alignat*}{10}
%    & F : \Ty\,\diamond\ra\Con && F : \Tm\,\diamond\,(\Gamma\Ra\U) \cong \Tm\,(\diamond\ext\Gamma)\,\U \cong \Ty\,(\diamond\ext\Gamma) \\
%    & F\,\Gamma := \diamond\ext\Gamma && F\,A := \El\,(A[\p]\cdot\q) \\
%    & F : \Tm\,\diamond\,(\Delta\Ra\Gamma) \ra \Sub\,(\diamond\ext\Delta) (\diamond\ext\Gamma) && F : \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q))) \cong \Tm\,(\diamond\ext\Gamma)\,(\El\,(A[\p]\cdot\q)) \\
%    & F\,\gamma := (\p,\gamma[\p]\cdot\q) \hspace{3em} && F\,a := a[\p]\cdot\q
%  \end{alignat*}
  It is easy to check that $F$ preserves the CwF structure, i.e.\ it
  is a functor, $\epsilon : \Sub\,(F\,\diamond)\,\diamond$ is an
  isomorphism, $F\,(A[\gamma]) = F\,A[F\,\gamma]$, $F\,(a[\gamma]) =
  F\,a[F\,\gamma]$ and $(F\,\p,F\,\q) : \Sub\,(F\,(\Gamma\ext
  A))$ $(F\,\Gamma\ext F\,A)$ is an isomorphism. % hack
\end{proof}
The other round-trip, that is, starting with an SSC-model, termifying
it and comparing the result with the original SSC-model does not provide an
isomorphism because we can't define the map $F :
\Tm\,\diamond\,(\Delta\Ra\Gamma) \ra
\Sub\,(\diamond\ext\Delta)\,(\diamond\ext\Gamma)$ on
substitutions. Such a map is not definable only using single
substitutions (a $\Sub$ is either a lifted weaking or a lifted single substitution, but we don't know anything about the relationship between $\Delta$ and $\Gamma$). However, the set of terms and types are still
isomorphic after the round-trip. We leave formulating the right notion
of contextual isomorphism for SSCs as future work.

>>>>

\title{Type theory with single substitutions\thanks{This research was supported by the HOTT ERC Grant 101170308.}}
\author{Ambrus Kaposi
\institute{Eötvös Loránd University (ELTE)\\
Budapest, Hungary}
\email{akaposi@inf.elte.hu}
\and
Szumi Xie
\institute{Eötvös Loránd University (ELTE)\\
Budapest, Hungary}
\email{szumi@inf.elte.hu}
}
\def\titlerunning{Type theory with single substitutions}
\def\authorrunning{A. Kaposi \& Sz. Xie}

@@

What is type theory? Here we refer to type theory as a particular
formal system based on Martin-Löf's original definition
\cite{MARTINLOF197573}, and not to the study of type systems
(e.g.\ \cite{DBLP:books/daglib/0005958}).

@@

However, even at the GAT level of abstraction, there are some choices
to be made about the instantiation operation. It is usually convenient
to use parallel substitutions, which means that the instantiation
operation replaces all variables in a context by terms in another
context \emph{at the same time}, that is, $t[x_0\mapsto
  u_0,\dots,x_n\mapsto u_n]$ where $x_0,\dots,x_n$ are the variables
in the context of $t$. We call $(x_0\mapsto u_0,\dots,x_n\mapsto u_n)$
a parallel substitution, where the terms $u_0,\dots,u_n$ are
all in the same context. Parallel substitutions are very natural in
the algebraic setting because they form the morphisms in a category
where objects are contexts (lists of types), then types/terms are
presheaves, context extension is a representability condition, so the
equations of the algebraic theory are forced by the categorical
structure. Some examples of algebraic definitions of type theory using
parallel substitutions ranging from the more syntactic to the more
semantic: Ehrhard's notion \cite{ehrhard,coquandEhrhard}, categories
with families (CwF \cite{DBLP:conf/types/Dybjer95,Castellan2021}),
natural models \cite{DBLP:journals/mscs/Awodey18}, contextual
categories \cite{DBLP:journals/apal/Cartmell86}, C-systems
\cite{DBLP:journals/lmcs/AhrensLV18}, locally cartesian closed
categories \cite{DBLP:journals/mscs/ClairambaultD14}, path categories
\cite{DBLP:journals/tocl/Berg18}.

@@

\vspace{-1em}
\subparagraph*{Formalisations of type theory.}
Most computer formalisations of type theory are extrinsic
\cite{DBLP:journals/pacmpl/0001OV18,DBLP:conf/cpp/AdjedjLMPP24,DBLP:journals/jar/SozeauABCFKMTW20}:
they define the syntax as abstract syntax trees, and equip it with
typing and conversion relations. A higher level representation is
working in setoid hell \cite{chapman09eatitself,setoidhell}: terms are
intrinsically well-typed, but conversion is still an explicit
relation. For formalising the GAT-level syntax, one needs a stronger
metatheory than plain Agda or Coq: the metatheory has to support
quotient inductive-inductive types (QIITs
\cite{DBLP:journals/pacmpl/KaposiKA19}), in other words, initial
models of GATs. Altenkirch and Kaposi
\cite{DBLP:conf/popl/AltenkirchK16} formalised the syntax of type
theory using postulated QIITs in Agda, together with parametricity and
normalisation \cite{lmcs:4005} proofs. Brunerie and de Boer
\cite{initiality-agda} constructed the initial contextual category in
Agda using postulated quotients. Although we don't have a general
proof, we have experimental evidence that QIITs are supported by the
cubical set model \cite{DBLP:conf/lics/CoquandHM18} and the setoid
model \cite{kaposi-qiit-setoid}, thus Cubical Agda
\cite{DBLP:journals/jfp/VezzosiMA21} and setoid/observational type
theory \cite{DBLP:conf/mpc/AltenkirchBKT19,DBLP:phd/hal/Pujet22} support QIITs. For example,
in Cubical Agda, set-truncated and groupoid-truncated syntaxes of type
theory have been formalised \cite{cohtt}. Without QIITs, tricks such
as shallowly embedding the syntax can be used to formalise metatheoretic
results about type theory such as gluing
\cite{glue}, and its special cases
canonicity and parametricity \cite{DBLP:conf/mpc/KaposiKK19}.

@@

First of all, we need a sort of terms which have to be indexed by
types because we only want well-typed terms. Both types and terms can
include variables, and to keep track of the currently available
variables, we also index them by contexts: a context is a list of
types, the length is the number of available variables, and we add new
variables at the end (that is, they are snoc-lists rather than
cons-lists). Types are also indexed by their universe level, this is a
technical requirement for avoiding Russell's paradox. The index $i$ is
an implicit argument of $\Tm$.
\begin{equation*}
\Con : \Set \hspace{3em} 
\Ty : \Con\ra\N\ra\Set \hspace{3em} 
\Tm : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set
\end{equation*}
Just as lists have two constructors, there are two ways to form
contexts: the empty context $\diamond$ and context extension $\ext$
which is like the snoc operation for lists. Context extension takes a
type which can have variables in the context preceding the type.
\[
\diamond : \Con \hspace{3em}
\blank\ext\blank : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con
\]
We will refer to variables by their distance from the end of the
context. We would like to describe the operation providing the last
variable in a context, first of all we need an operation providing the last
variable in the context. This is the zero De Bruijn index \cite{DEBRUIJN1972381}, which we
denote by $\q : \Tm\,(\Gamma\ext A)\,A'$. It is a term in an
extended context, and it should have type $A$, but the issue is that
$A : \Ty\,\Gamma$, and $A' : \Ty\,(\Gamma\ext A)$. We need a way to
weaken the type $A$ to obtain such an $A'$. For this, we will
introduce an operation $\blank[\p] : \Ty\,\Gamma\,i\ra\Ty\,(\Gamma\ext
A)\,i$ and define $A' := A[\p]$. Instead of introducing just
$\blank[\p]$, we generalise a bit and add a new sort $\Sub$. For
now, $\Sub$ only has the single element $\p$ and we introduce the operation
$\blank[\blank]$ called \emph{instantiation}. Elements of $\Sub$
will later be called substitutions, hence the name, but for now, we
only have single end-of context weakenings in $\Sub$. The context
$\Delta$ is called the domain, $\Gamma$ the codomain of a
$\Sub\,\Delta\,\Gamma$.
\begin{alignat*}{10}
  & \Sub && : \Con\ra\Con\ra\Set &&\hspace{1.5em}
  & \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \hspace{1.5em}
  \p && : \Sub\,(\Gamma\ext A)\,\Gamma \hspace{1.5em}
  && \q && : \Tm\,(\Gamma\ext A)\,(A[\p])
\end{alignat*}
The operations $\p$ and $\q$ take three implicit parameters, $\Gamma$,
$i$ and $A$, while $\blank[\blank]$ takes $\Gamma$, $i$ and $\Delta$
implicitly. Still, we only have the last variable $\q$ in the context,
we don't have e.g.\ the last but one variable in $\Tm\,(\Gamma\ext
A\ext B)\,(A[\p][\p])$ where we had to weaken $A$ twice to make it fit
with its context. To obtain more variables, we also allow weakening of
terms, more precisely, we introduce an instantiation operation for
terms with an overloaded notation.
\begin{alignat*}{10}
& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma])
\end{alignat*}
Note that this is a dependent function as the type of the resulting
term has to be weakened the same way as the term itself. Now we can
define all variables counting from the end of the context (De Bruijn
indices): $0 := \q$, $1 := \q[\p]$, $2 := \q[\p][\p]$,
$3 := \q[\p][\p][\p]$, and so on.

@@

\cite{DBLP:conf/mpc/KaposiKK19} we call it termification, as most sorts in the

@@

B-systems are B-frames
together with substitution, weakening and generic element
operations. A B-frame is given by sets $B_i$, $\tilde{B}_i$ and
functions between them as shown below. \\
\begin{tikzpicture}
  \node (b0) at (4,0) {$\top$};
  \node (b1) at (5,0) {$B_1$};
  \node (b2) at (12,0) {$B_2$};
  \node (b3) at (15,0) {$\dots$};
  \node (c1) at (7,1) {$\tilde{B}_1$};
  \node (c2) at (14,1) {$\tilde{B}_2$};
  \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{ft}_0$} (b0);
  \draw[->,font=\scriptsize] (b2) edge node[below] {$\mathsf{ft}_1$} (b1);
  \draw[->,font=\scriptsize] (b3) edge node[below] {$\mathsf{ft}_2$} (b2);
  \draw[->,font=\scriptsize] (c1) edge node[above] {$\partial_1$} (b1);
  \draw[->,font=\scriptsize] (c2) edge node[above] {$\partial_2$} (b2);
\end{tikzpicture} \\
Using our notation, this is the following data ($\diamond$ is the empty context, $\ext$ is context extension). \\
\begin{tikzpicture}
  \node (b0) at (4,0) {$\top$};
  \node (b1) at (5,0) {$\Ty\,\diamond$};
  \node (b2) at (12,0) {$(A:\Ty\,\diamond)\times\Ty\,(\diamond\ext A)$};
  \node (b3) at (15,0) {$\dots$};
  \node (c1) at (7,1)  {$(A:\Ty\,\diamond)\times\Tm\,\diamond\,A$};
  \node (c2) at (14,1) {$\big((A:\Ty\,\diamond)\times(B:\Ty\,(\diamond\ext A))\big)\times\Tm\,(\diamond\ext A)\,B$};
  \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{fst}$} (b0);
  \draw[->,font=\scriptsize] (b2) edge node[below] {$\mathsf{fst}$} (b1);
  \draw[->,font=\scriptsize] (b3) edge node[below] {$\mathsf{fst}$} (b2);
  \draw[->,font=\scriptsize] (c1) edge node[above] {$\mathsf{fst}$} (b1);
  \draw[->,font=\scriptsize] (c2) edge node[above] {$\mathsf{fst}$} (b2);
\end{tikzpicture} \\
The substitution operation $\mathbb{S}$ for an $x:\tilde{B}_{n+1}$ is
a homomorphism of the slice B-frames $\mathbb{B}/\partial(x) \ra
\mathbb{B}/\mathsf{ft}(\partial(x))$. In our notation, $x = (\Gamma,
A, a)$ where $a : \Tm\,\Gamma\,A$, and $\mathbb{S}$ corresponds to the
following functions for $\Ty$ (a map between the bottom rows of the
diagrams):
\begin{alignat*}{10}
  & \blank[\langle a\rangle] && : \Ty\,(\Gamma\ext A)\ra\Ty\,\Gamma \\
  & \blank[\langle a\rangle^+] && : \Ty\,(\Gamma\ext A\ext B)\ra\Ty\,(\Gamma\ext B[\langle a\rangle]) \\
  & \blank[\langle a\rangle^{++}] && : \Ty\,(\Gamma\ext A\ext B\ext C)\ra\Ty\,(\Gamma\ext B[\langle a\rangle]\ext C[\langle a\rangle^+]) \\
  & \dots,
\end{alignat*}
and similarly $\mathbb{S}$ also includes all the (lifted) substitution
operations for terms (top rows of the
diagrams). Analogously, the weakening operation corresponds to
$\blank[\p^{+\dots+}]$ operations, and the generic element is $\q$ in
our notation. The six groups of equations correspond to our
$[\langle\rangle][]$, $[\p][^+]$, $\q[^+]$, $[\p][\langle\rangle]$,
$\q[\langle\rangle]$, $[\p^+][\langle\q\rangle]$ equations, in this
order. However we don't include the lifted versions of these equations,
equations $[\langle\rangle][]$, $[\p^+][\langle\q\rangle]$ are only
stated for types, and $\q[^+]$, $\q[\langle\rangle]$ are only stated
for terms. The missing equations are admissible, see Section
\ref{sec:admissible}. In the presence of $\U$ and $\Pi$, we
reduce the needed equations even more, see Section
\ref{sec:minimisation}.

@@

We show that the roundtrip CwF $\longrightarrow$ SSC
$\xrightarrow{\text{termification}}$ CwF results in a contextually
isomorphic CwF. A contextual isomorphism
\cite{DBLP:journals/corr/abs-2211-07487} is a weak CwF-morphism
(pseudomorphism, i.e.\ context extension and the empty context are
only preserved up to isomorphism) which is bijective on types and
terms. Note that a contextual isomorphism preserves all type formers
specified by universal properties.
\begin{problem}
  Given a CwF (with type formers) $M$, let $M'$ denote the CwF
  obtained by first seeing it as an SSC model and then termifying it. We
  construct a contextual isomorphism between $M'$ and $M$.
\end{problem}
\begin{proof}[Construction]
  We denote the components of the contextual isomorphism $F$ as follows.
  \begin{alignat*}{10}
    & F : \Con_{M'}\ra\Con_M && F : \Ty_{M'}\,\Gamma \cong \Ty_M\,(F\,\Gamma) \\
    & F : \Sub_{M'}\,\Delta\,\Gamma\ra\Sub_M\,(F\,\Delta)\,(F\,\Gamma)\hspace{3em} && F : \Tm_{M'}\,\Gamma\,A \cong \Tm_M\,(F\,\Gamma)\,(F\,A)
  \end{alignat*}
  We define them in the same order, omitting the $M$ subscripts:
  $F\,\Gamma := \diamond\ext\Gamma$, 
  $F\,\gamma := (\p,\gamma[\p]\cdot\q)$, 
  $F\,A := \El\,(A[\p]\cdot\q)$, 
  $F\,a := a[\p]\cdot\q$.
%  \begin{alignat*}{10}
%    & F : \Ty\,\diamond\ra\Con && F : \Tm\,\diamond\,(\Gamma\Ra\U) \cong \Tm\,(\diamond\ext\Gamma)\,\U \cong \Ty\,(\diamond\ext\Gamma) \\
%    & F\,\Gamma := \diamond\ext\Gamma && F\,A := \El\,(A[\p]\cdot\q) \\
%    & F : \Tm\,\diamond\,(\Delta\Ra\Gamma) \ra \Sub\,(\diamond\ext\Delta) (\diamond\ext\Gamma) && F : \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q))) \cong \Tm\,(\diamond\ext\Gamma)\,(\El\,(A[\p]\cdot\q)) \\
%    & F\,\gamma := (\p,\gamma[\p]\cdot\q) \hspace{3em} && F\,a := a[\p]\cdot\q
%  \end{alignat*}
  It is easy to check that $F$ preserves the CwF structure, i.e.\ it
  is a functor, $\epsilon : \Sub\,(F\,\diamond)\,\diamond$ is an
  isomorphism, $F\,(A[\gamma]) = F\,A[F\,\gamma]$, $F\,(a[\gamma]) =
  F\,a[F\,\gamma]$ and $(F\,\p,F\,\q) : \Sub\,(F\,(\Gamma\ext
  A))$ $(F\,\Gamma\ext F\,A)$ is an isomorphism. % hack
\end{proof}
The other round-trip, that is, starting with an SSC-model, termifying
it and comparing the result with the original SSC-model does not provide an
isomorphism because we can't define the map $F :
\Tm\,\diamond\,(\Delta\Ra\Gamma) \ra
\Sub\,(\diamond\ext\Delta)\,(\diamond\ext\Gamma)$ on
substitutions. Such a map is not definable only using single
substitutions (a $\Sub$ is either a lifted weakening or a lifted single substitution, but we don't know anything about the relationship between $\Delta$ and $\Gamma$). However, the set of terms and types are still
isomorphic after the round-trip. We leave formulating the right notion
of contextual isomorphism for SSCs as future work.
